SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-11-19 2017-Xudong-SMT/ActaInformatica.tex
Acta Inf. paper: Typos.

\ERIC{This shows an example of the generic ``FUN'' constructor used to construct expressions (of the Action sort, or of any data sort) in any algebra presentations.
  Here \texttt{FUN} will be instanciated e.g. as, $\delta(x)$ in the form of $\texttt{FUN}_{Action, Data}(\delta, x)$ where $x$ is a variable of type $Data$.
  %Here we have $arity(\texttt{FUN}_{Action, Data})=2$ and $\texttt{FUN}_{Action, Data}: Action, Data\to Action$.
  Note that the same action constructor $\delta$ is also used as a constant action in the controller $CState$.
}

@@

\begin{example}[pNet of the ``Enable'' Operator]\label{example:pNetEnable}
\ERIC{
  Consider the pNet $EnableState$ shown in Fig. \ref{schema:enable-pnets}, upper part. It has two holes $P$ and $Q$, and one PLTS $CState$. Note that our graphical syntax presents pNet nodes and synchronisation vectors as finite arrays, where the index sets follow the positions in the arrays.
  In this example we use \emph{synchronised actions}, to express actions that cannot be further synchronised, similar to the silent action $\tau$ of CCS. We denote them as any action expression but with the text underlined,  e.g. $\underline{\delta(x_1)}$. Such synchronised actions do not play any special role for defining the semantics of pNets, but as one can expect, they will be crucial for defining weak equivalences. The pNet can be given as follow:
\begin{itemize}
	\item[$\bullet$] $\pNet=\{CState\}$;
	\item[$\bullet$] $J=\{P, Q\}$;
	\item[$\bullet$] $\symb{SV}=$ as listed on the right side of each pNet node.
        %%   \{\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
	%% \langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
	%% \langle \text{-}, a_2, r\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}$.
\end{itemize}
Consider the first  vector of $EnableState$, written as $\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y_1. a_1\neq \delta(y_1)]$. It denotes a possible behavior of $P>>Q$, for any possible action of the hole $P$ that respects the vector predicate, that is, any action that does not match $\delta(y_1)$, for any value of the variable $y_1$.
Remark that the data-oriented encoding in pNet $EnableData$ has the same synchronisation vectors, up to the usual alpha conversion. Only the controller $CState$ has been replaced by $CData$.
}
\end{example}

@@

  \label{OT:example-2}
        \begin{mathpar}
	  OT_2'  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x_2)}_P ~~
          \xrightarrow{acc(x_2)}_Q \} ~~
          [v_1=0 \land v_2 = 0 \land \forall y_1. acc(x_2) \neq
            \delta(y_1)] ~~ \{v_1:=1\}
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x_2)}} \ostate{0,0}}
	\end{mathpar}
        where $v_1$ and $v_2$ are respectively the state variables of
        the first and second $EnableData$ pNets. Indeed the global
        state space of this version is reduced to $\ostate{0,0}$, and
        the state change is replaced by the assignment ``$v_1:=1$''.

@@

\QIN{The matching would also fail if the active actions in the vector and the active behaviours in the chosen combination are not matched by pattern-matching or unification. Such case will be checked later, together with the guards collected from synchronisation vectors and from PLTS transtions, using the satisfiability check in the SMT engine.}

@@

\ERIC{Each call to the SMT engine will return either SAT or UNSAT results, or may fail due to time bounds. Also, the accuracy of a `SAT' result may depend on the characteristics of the specific theory used to axiomatise the data domains and operators. Such theories may be intrinsically undecidable, or may not  be properly handled by the SMT engine (e.g. for recursive datatypes).
However, while building the open automaton, which is a symbolic representation of the pNet behavior, it is not logically incorrect to keep potentially unsatisfiable transitions, so we will treat undecidable results, and time bounds failures, as SAT. So the open automaton may have more transitions and states than the theoretical minimum.}

@@

Here the behavior of the hole ``$P$'' ($a_1$ in Example~\ref{schema:unsat-ot}), after matching with the synchronisation vectors, appears as a variable
``|a1:sva\_SVA0:15:1|'', and ``$R$'' ($acc(x_2)$) as ``|(ACT delta
|x:sva\_SV4:1:2|''. Line 16 requires that $a_1 = \delta(x_2)$ to match
the second synchronisation vector of the root pNet node, and lines
17-18 that $a_1$ does not match $\delta(x_1)$ for any $x_1$. 
We also display  the diagnosis (``SAT'' or ``UNSAT'') generated by Z3.

@@

The open automata generated for the two encodings of Enable in
Figure~\ref{schema:enable-pnets} are displayed in
Figure~\ref{schema:enable-OAs}. 
\QIN{The initial transition of $EnableData$ is transformed to the initialisation of the state variables, which is $\{v\leftarrow 0\}$, for its open automata.}
Note that although these automata look very similar to the PLTS
controllers from Figure~\ref{schema:enable-pnets}, their nature is very
different: they represent the behavior of the whole system, including
the holes $P$ and $Q$. Their transitions do not rely on a particular
implementation of the controllers, but only encode the 
relations between the actions of the holes.
We have shown in \cite{henrio:Forte2016} 
\YX{that the two open automata are equivalent with respect to a notion of (symbolic) FH-bisimulation.}
%how to prove that they are bisimilar.

@@

\ERIC{The original specification of the nanosatellite software, using BIP architecture patterns, was handling data in the BIP components behavior, but was not able to use data in the synchronisation between components. We have shown in \cite{BHM19:Coordination} how to overcome this limit, allowing data constraints in BIP architecture connectors, and providing an encoding of such architectures in terms of open pNets.}

>>>>

\ERIC{This shows an example of the generic $\texttt{FUN}$ constructor used to construct expressions (of the Action sort, or of any data sort) in any algebra presentations.
  Here \texttt{FUN} will be instanciated e.g. as, $\delta(x)$ in the form of $\texttt{FUN}_{Action, Data}(\delta, x)$ where $x$ is a variable of type $Data$.
  %Here we have $arity(\texttt{FUN}_{Action, Data})=2$ and $\texttt{FUN}_{Action, Data}: Action, Data\to Action$.
  Note that the same action constructor $\delta$ is also used as a constant action in the controller $CState$.
}

@@

\begin{example}[pNet of the ``Enable'' Operator]\label{example:pNetEnable}
\ERIC{
  Consider the pNet $EnableState$ shown in Fig. \ref{schema:enable-pnets}, upper part. It has two holes $P$ and $Q$, and one PLTS $CState$. Note that our graphical syntax presents pNet nodes and synchronisation vectors as finite arrays, where the index sets follow the positions in the arrays.
  In this example we use \emph{synchronised actions}, to express actions that cannot be further synchronised, similar to the silent action $\tau$ of CCS. We denote them as any action expression but with the text underlined,  e.g. $\underline{\delta(x_1)}$. Such synchronised actions do not play any special role for defining the semantics of pNets, but as one can expect, they will be crucial for defining weak equivalences. The pNet can be given as follow:
\begin{itemize}
	\item[$\bullet$] $\pNet=\{CState\}$;
	\item[$\bullet$] $J=\{P, Q\}$;
	\item[$\bullet$] $\symb{SV}=$ as listed on the right side of each pNet node.
        %%   \{\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
	%% \langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
	%% \langle \text{-}, a_2, r\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}$.
\end{itemize}
Consider the first  vector of $EnableState$, written as $\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y_1. a_1\neq \delta(y_1)]$. It denotes a possible behaviour of $P>>Q$, for any possible action of the hole $P$ that respects the vector predicate, that is, any action that does not match $\delta(y_1)$, for any value of the variable $y_1$.
Remark that the data-oriented encoding in pNet $EnableData$ has the same synchronisation vectors, up to the usual alpha conversion. Only the controller $CState$ has been replaced by $CData$.
}
\end{example}

@@

  \label{OT:example-2}
        \begin{mathpar}
	  OT_2'  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x_2)}_P ~~
          \xrightarrow{acc(x_2)}_Q \} ~~
          [v_1=0 \land v_2 = 0 \land \forall y_1. acc(x_2) \neq
            \delta(y_1)] ~~ \{v_1\leftarrow1\}
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x_2)}} \ostate{0,0}}
	\end{mathpar}
        where $v_1$ and $v_2$ are respectively the state variables of
        the first and second $EnableData$ pNets. Indeed the global
        state space of this version is reduced to $\ostate{0,0}$, and
        the state change is replaced by the assignment ``$v_1\leftarrow1$''.

@@

\QIN{The matching would also fail if the active actions in the vector and the active behaviours in the chosen combination are not matched by pattern-matching or unification. Such case will be checked later, together with the guards collected from synchronisation vectors and PLTS transitions, using the satisfiability check in the SMT engine.}

@@

\ERIC{Each call to the SMT engine will return either SAT or UNSAT results, or may fail due to time bounds. Also, the accuracy of a `SAT' result may depend on the characteristics of the specific theory used to axiomatise the data domains and operators. Such theories may be intrinsically undecidable, or may not  be properly handled by the SMT engine (e.g. for recursive datatypes).
However, while building the open automaton, which is a symbolic representation of the pNet behaviour, it is not logically incorrect to keep potentially unsatisfiable transitions, so we will treat undecidable results, and time bounds failures, as SAT. So the open automaton may have more transitions and states than the theoretical minimum.}

@@

Here the behaviour of the hole ``$P$'' ($a_1$ in Example~\ref{schema:unsat-ot}), after matching with the synchronisation vectors, appears as a variable
``|a1:sva\_SVA0:15:1|'', and ``$R$'' ($acc(x_2)$) as ``|(ACT delta
|x:sva\_SV4:1:2|''. Line 16 requires that $a_1 = \delta(x_2)$ to match
the second synchronisation vector of the root pNet node, and lines
17-18 that $a_1$ does not match $\delta(x_1)$ for any $x_1$. 
We also display  the diagnosis (``SAT'' or ``UNSAT'') generated by Z3.

@@

The open automata generated for the two encodings of Enable in
Figure~\ref{schema:enable-pnets} are displayed in
Figure~\ref{schema:enable-OAs}. 
\QIN{The initial transition of $EnableData$ is transformed to the initialisation of the state variables, which is $\{v\leftarrow 0\}$, for its open automata.}
Note that although these automata look very similar to the PLTS
controllers from Figure~\ref{schema:enable-pnets}, their nature is very
different: they represent the behaviour of the whole system, including
the holes $P$ and $Q$. Their transitions do not rely on a particular
implementation of the controllers, but only encode the 
relations between the actions of the holes.
We have shown in \cite{henrio:Forte2016} 
\YX{that the two open automata are equivalent with respect to a notion of (symbolic) FH-bisimulation.}
%how to prove that they are bisimilar.

@@

\ERIC{The original specification of the nanosatellite software, using BIP architecture patterns, was handling data in the BIP components behaviour, but was not able to use data in the synchronisation between components. We have shown in \cite{BHM19:Coordination} how to overcome this limit, allowing data constraints in BIP architecture connectors, and providing an encoding of such architectures in terms of open pNets.}
